---
title: "ESM 206 Lab 5"
author: "Nicole Schmidt"
date: "10/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Lab 5 objectives

- Learn to parse dates with 'lubridate' package
- Get counts of observations with count()
- Uncount() rows 
- One and two-sample t-test
- Create a heatmap with geom_tile()

```{r, include = FALSE}
#Read in data and attach packages
library(tidyverse)
library(here)
library(janitor)
#use here package here function, look at folder (data) than file (csv)
lobster_abundance <- read_csv(here::here("data", "lobster_abundance.csv"), 
                              na = "-99999") %>% 
  janitor::clean_names() 
#getting data into tidy format
#replace -99999 values with na values
  

#looking at the data table we want r to not understand these as a character for data
```

Use tidyr::uncount() function to convert our lobster data from frequency format to case format

```{r}
lobster_tidy <- lobster_abundance %>% 
  tidyr::uncount(lobster_count)

#More tidying of data 
#give uncount the column that contains the untidy info 
#every individual lobster is given its own row ie. every observation has own row
#frequency to case format

```

#### Exploratory data visualization 

were only going to consider 'site' as our varianle of interest

```{r}

ggplot(lobster_tidy, aes(x = site,
                         y = size_mm))+
  geom_jitter(aes(color = site,
                  alpha = 0.5,
                  width = 0.2))

#so many obersvations cant see what this distribution really looks like
# ask questions from data? where is distribution of mean, median mode? jitter plots okay but not for when you have a ton of data

#histogram

ggplot(lobster_tidy, aes(x = size_mm)) +
  geom_histogram(aes(fill = site))+ #just get stacked histogram which is confusing so
  facet_wrap(~site, scales = "free") # creates multiple graphs for your data for different sites

#look at graph and ask what does the general shapt look like? overall bell shaped, satisfy an assumptuon of normality. dont want t look at things in one way. historgrams will change based on how many bins there are. So lets look at a different visualization of this data

#using quantile quantile plots to check for what the distribution of the data is. Is it linear? linear relationship 
ggplot(lobster_tidy, aes(sample = size_mm))+
  geom_qq()+#distribution for all observations = linear observations are normally distrubuted.
  facet_wrap(~site) #check for linear relationship (normally distributed) for each site 

```

#### Convert the 'date' column to class 'Date'

we'll use the lubridate package to convert to Date format and then to help us easily parse month & year 

```{r}
#existing structre for dates is in month/day/year we want to use lubridate to convert date format into 

lobster_date <- lobster_tidy %>% 
  mutate(date_new = lubridate::mdy(date)) #new name is date_new change to month day year

```

Now lets parse year and month using lubridate::month() and lubridate::(year)
```{r}

#lubridate package makes it easier to work with dates and times. If collecting inputting date put in non ambiguous format. Make sure way you are reading in data makes sense. lubridate great for this

lobster_parse_date <- lobster_date %>% 
  mutate(obs_month = lubridate::month(date_new, label = TRUE),
         obs_year = lubridate::year(date_new))


# add new column for month, label = true will change month to month name and know that it is an ordered factor and the order should be mantained. 

```

Now lets find counts of observed lobsters based on different hierarchical groupings: use count function. groups, counts, ungroups. 

First lets count lobsters by year and month
```{r}

lobster_ym <- lobster_parse_date %>% 
  dplyr::count(obs_year, obs_month)

lobster_y <- lobster_parse_date %>% 
  dplyr::count(obs_year)

lobster_site <- lobster_parse_date %>% 
  dplyr::count(site)

```

If we want to create a summary table that contains statistics other than counts by group, it's easier to use use group_by() + n()

```{r}

#good for using multiple statistics when creating summary table

lobster_summary <- lobster_parse_date %>% 
  group_by(site) %>% 
  summarize(
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE),
    lobster_number = n())

lobster_summary

```

#### Find confidence intervals 

use t.test() function to find confidence intervals for one sample and perform t tests to compare means of two samples (...this will be covered conceptually in lectures for week 6)

```{r}
ivee_lobster <- lobster_tidy %>% 
  filter(site == "IVEE") %>% 
  pull(size_mm)

t.test(ivee_lobster)

```

#### Two- sample t-test to compare means 

I want to know if there is a significant difference between lobsters at different sites?
null hypotheses- samples taken from populations with the same mean
If draw 2 perfectly random samples from the same population the mean the difference in their means would be 0. but thats impossible bc there is always error. 

Are my 2 means so different that i have sufficient evidence that they were drawn from population samples with 2 different means??

We want to ask if there is a significant difference in lobster lengths at Naples and Mohawk reefs?

We've done our necessary exploratory analysis to determine that a 2 sample t-test for means comparison is appropriate.

```{r}

naple_sample <- lobster_tidy %>% 
  filter(site == "NAPL") %>% 
  pull(size_mm)

mohk_sample <- lobster_tidy %>% 
  filter(site == "MOHK") %>% 
  pull(size_mm)

mn_ttest <- t.test(mohk_sample, naple_sample)
mn_ttest
```

There is a significant difference in lobster lengths between Naples and Mohawk Reef. 
(t(df) = statistice, p < 0.001, alpha = 0.05).

What are the actual means? 
Whats the effect size?
Whats the structure of the data? 

```{r}

lobster_mn <- lobster_tidy %>% 
  filter(site %in% c("NAPL", "MOHK"))

mn_ttest2 <- t.test(size_mm ~ site, data = lobster_mn)
mn_ttest2

mn_ttest2$p.value
mn_ttest2$statistic

```

Here is the t-statistic: 'r mn_ttest2$p.value'

Here is my t-statistic: 'r mn_ttest2$statistic'

#### Now a heat map

```{r}

lobster_ys <- lobster_parse_date %>% 
  count(obs_year, site)

ggplot(data = lobster_ys, aes(x = obs_year, y = site))+
  geom_tile(aes(fill = n))

```


